---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: Chemical Effects on Mammals
subtitle: https://github.com/c-reents/DataAnalyticsProject
author: Caroline Reents
abstract: "Experimental overview. This section should be no longer than 250 words."
fontsize: 12pt
mainfont: Times New Roman
fig_width: 6 
fig_height: 4 
---

<Information in these brackets are used for annotating the RMarkdown file. They will not appear in the final version of the PDF document>

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

<Setup the global options for the R chunks in your document>

<Note: set up autoreferencing for figures and tables in your document>

```{r, setup, echo=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(readr)
library(viridis)
library(RColorBrewer)
library(kableExtra)





# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)


#upload the data set
Mammal_dat_RAW <- read.csv("ECOTOX_Mammals_RAW.csv")
View(Mammal_dat_RAW)

#setting up autoreferencing 

```


# Research Question and Rationale

<Paragraph detailing the rationale for your analysis. What is the significant application and/or interest in this topic? Connect to environmental topic(s)/challenge(s).>

<Paragraph detailing your research question(s) and goals. What do you want to find out? Include a sentence (or a few) on the dataset you are using to answer this question - just enough to give your reader an idea of where you are going with the analysis.>

This data set includes the effects of certain pharmaceutical or health care drugs or chemicals on the mortality of various mammals. Each data point is related to a specific study done on that chemical with a certain mammal. Mortality is measured in various "endpoints" which refer to how much mortality was seen in each study. There are 24 different chemicals in this data set, large groups of these chemicals are either cancer-combating drugs or antibiotics. For this project I aim to determine if different mammals are more suscepitble to death from being exposed to these chemicals, or if mortality is highly dependent on concentration of chemical. I also want to determine if the cancer drugs and the antibiotics differ in their likelihood to cause death. 

\newpage

# Dataset Information

<Information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the README file for the dataset but formatted in a way that is more narrative.>

The following is a list of all chemicals tested in this data set pulled from the ECOTOX webpage. Each data point is the result of a study, so each data point seen on visualizations represents the findings of one or a group of researchers published study. 

The data was pulled filtering for mammals, effects of mortality, all endpoints, and chemical subgroup "pharmaceutical personal care products". The drugs tested in this data set were: 

- 17alpha-ethynylestradiol : used in BC
- 6-mercaptopurine : used for cancer and autoimmune
- Ampicillin - antibiotic
- Betamethasone - steroid
- Clindamyacin - antibiotic
- Dexamethasone - steroid
- Dexamethasone sodium - steroid
- Diazepam - pain killer
- Dimenhydrinate - motion sickness
- Erythomycin - antibiotic
- Gentamycin - antibiotic
- Lindane - treats scabies
- Methimazole - treats hyperthyroid
- Methotrexate - cancer autoimmune treatment 
- Metronidazole - antibiotic
- Phenobarbital - sedative
- Prednisone - steroid
- Propranolol - beta-blocker
- Tetracycline - antibiotic
- Tetracycline hydrochloride - antibiotic
- Thalidomide - cancer treeatment
- Trans-Retinoic acid - cancer
- Triamcinolone - treats inflammation
- Warfarin - prevents blood clots  

It is important to note that the endpoints were defined as follows: 

* EC10, EC50: effective concentration to x % of test organisms
* ET50: effective response time to 50 % of test organisms
* LC0 through LC95: lethal concentration to x % of test organisms
* LD50, LD95, LD99: lethal dose to x % of test organisms
* LOEC: lowest observable effect concentration
* LOEL: lowest observable effect level
* LT50: time to 50 % mortality of organisms
* NOEC: no observable effect concentration
* NOEL: no observable effect level
* NR: not reported
* NR-LETH: 100 % mortality or 0 % survival of test organisms
* NR-ZERO: 0 % mortality or 100 % survival of test organisms

<Add a table that summarizes your data structure. This table can be made in markdown text or inserted as a `kable` function in an R chunk. If the latter, do not include the code used to generate your table.>
```{r mammal_table, echo=FALSE}
mammal_table<-kable(Mammal_dat_RAW)%>%
  kable_styling(bootstrap_options = "striped", full_width = F)

print(mammal_table)



```


\newpage

# Exploratory Data Analysis and Wrangling

<Include R chunks for 5+ lines of summary code (display code and output), 3+ exploratory graphs (display graphs only), and any wrangling you do to your dataset(s).> 
```{r summary}
#summary code
colnames(Mammal_dat_RAW)
class(Mammal_dat_RAW$Pub..Year)
summary(Mammal_dat_RAW$Common.Name)

summary(Mammal_dat_RAW$Chemical.Name)
class(Mammal_dat_RAW$Chemical.Name)
#Thalidomide and warfarin have most data points

class(Mammal_dat_RAW$Conc..Mean..Std.)
Mammal_dat_RAW$Conc..Mean..Std.<- as.integer(Mammal_dat_RAW$Conc..Mean..Std.)
#changed cancentration to an integer rather than numeric

summary(Mammal_dat_RAW$Pub..Year)
Mammal_dat_RAW$Pub..Year<-as.factor(Mammal_dat_RAW$Pub..Year)
summary(Mammal_dat_RAW$Pub..Year)
#treated this as a factor so that I could see how many studies took place in each year

summary(Mammal_dat_RAW$Endpoint)
class(Mammal_dat_RAW$Endpoint)
```


```{r wrangling}
ECOTOX_LethZero <- Mammal_dat_RAW %>%
  filter(Endpoint == "NR-LETH" | Endpoint =="NR-ZERO")

ECOTOX_Lethal <- Mammal_dat_RAW %>%
  filter(Endpoint == "NR-LETH")



class(ECOTOX_LethZero$Endpoint)
ECOTOX_LethZero$Endpoint<-as.character(ECOTOX_LethZero$Endpoint)

summary(Mammal_dat_RAW$Chemical.Name)

ECOTOX_antibiotics<- Mammal_dat_RAW %>%
  filter(Chemical.Name == "Ampicillin"| 
           Chemical.Name == "Clindamycin"| 
           Chemical.Name =="Gentamycin"|
           Chemical.Name ==  "Metronidazole"|
           Chemical.Name == "Erythromycin"|
           Chemical.Name == "Tetracycline"|
           Chemical.Name == "Tetracycline hydrochloride")
View(ECOTOX_antibiotics)



#wrangle the data to only have the three most commonly studied organisms

Three_dat <- Mammal_dat_RAW %>%
  filter(Common.Name == "Afghan Pika" | Common.Name == "Norway Rat" | Common.Name =="House Mouse")

summary(Mammal_dat_RAW$Chemical.Name)

#wrangle data to combine endpoints categories = 0, 50, 100


Three_dat_mutateMortality<-mutate(Three_dat, Mortality.Rate = ifelse(Three_dat$Endpoint=="LD50", "Lethal", "Not Lethal"),
                                  ifelse(Three_dat$Endpoint=="LOEL", "Not Lethal", "Lethal"),
                                  ifelse(Three_dat$Endpoint=="NOEL", "Not Lethal", "Lethal"),
                                  ifelse(Three_dat$Endpoint=="NR-LETH", "Lethal", "Not Lethal"),
                                  ifelse(Three_dat$Endpoint=="NR-ZERO", "Not Lethal", "Lethal"))



ECOTOX_cancerdrugs<- Mammal_dat_RAW %>%
  filter(Chemical.Name == "6-Mercaptopurine"| 
           Chemical.Name == "Methotrexate"| 
           Chemical.Name =="Thalidomide"|
           Chemical.Name =="trans-Retinoic acid")



```


```{r Exploratory 1, echo=FALSE}
Explore1<-ggplot(Mammal_dat_RAW, aes(x = Chemical.Name, y = Common.Name)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Chemical Name", y="Species")+
  geom_point()
print(Explore1)
#E1cap <- fig_nums(name = "Explore1", caption = "Exploratory visualization 1, species tested for each chemcial")
```

```{r Exploratory 2, echo=FALSE}
Explore2<- ggplot(Mammal_dat_RAW, aes(x = Chemical.Name, y = Endpoint)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Chemical Name", y="Effect")+
  geom_point()
print(Explore2)
#E2cap <- fig_nums(name = "Explore2", caption = "Exploratory visualization 2, endpoint effects for each chemical")
```

```{r exploratory 3, echo=FALSE} 

Explore3<- ggplot(ECOTOX_Lethal, aes(x = Endpoint, y = Conc..Mean..Std.)) +
  ylim(0, 150)+
  labs(x="Effect on Mortality", y="Concentration")+
  geom_boxplot(aes(color=Chemical.Name))+
  theme(legend.position = "right")

  
print(Explore3)  
#E3cap <- fig_nums(name = "Explore3", caption = "Exploratory visualization 3, concentration of drugs assocaited with 100% mortality")
  
```

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, and the rationale for your approach.>

The first exploratory graph shows which drugs were tested on which animals. This helps visualize if each drug was tested on each animal, which could effect the results of cetain statistical tests. For the second exploratory figure, I wanted to see which drugs resulted in which endpoints. I was particulary interested to see which drugs resulted in 100% mortality (NR-LETH). For the third exploratory figure, I determined the spread of concentration for the drugs that caused 100% mortality (NR-LETH). I wanted to determine if certain drugs required higher concentrations to kill and if certain drugs killed at multiple concentrations.  



\newpage

# Analysis
<Include R chunks for 3+ statistical tests (display code and output) and 3+ final visualization graphs (display graphs only).>
```{r statistical tests} 

```

```{r visualization 1, echo=FALSE}
Visualization1<- ggplot(ECOTOX_antibiotics, aes(Chemical.Name, Endpoint)) +
                          labs(x="Chemical Name", y="Endpoint", 
                               color="Mean Concentration (mg/kg)", size="Number of Studies")+ 
                          scale_color_viridis(option="plasma")+
                          geom_count(aes(color=Conc..Mean..Std.))+
                          theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                          legend.position="right",legend.key.size= unit(0.5, "cm"),                            legend.title=element_text(size=10))
  

print(Visualization1)


#V1cap <- fig_nums(name = "Visual1", caption = "Antibiotics and their endpoints based on concentration. Larger dots indicate more studies.")

```

```{r visualization 2, echo=FALSE}
Visualization2<- ggplot(ECOTOX_cancerdrugs, aes(Chemical.Name, Endpoint)) +
                          labs(x="Chemical Name", y="Endpoint", 
                               color="Mean Concentration (mg/kg)", size="Number of Studies")+ 
                          scale_color_viridis(option="plasma")+
                          geom_count(aes(color=Conc..Mean..Std.), breaks=c(1,2,3,4,5))+
                          scale_size(breaks=c(1,2,3))+
                          theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                          legend.position="right",legend.key.size= unit(0.5, "cm"),                             legend.title=element_text(size=10))

print(Visualization2)

#V2cap <- fig_nums(name = "Visual2", caption = "Anti-cancer drugs and their endpoints based on concentration. Larger dots indicate more studies.")
```






```{r Visualization3, echo=FALSE}
Visualization3 <- ggplot(Three_dat_mutateMortality, aes(Mortality.Rate,))+
  geom_bar(stat = "count", position = "stack", aes(fill=Common.Name))+
  labs(x="Effect on Mortality", y="Number of Studies", fill="Species Exposed")+
  scale_fill_manual(values=c("orchid1", "khaki1", "turquoise2"))+
  theme(legend.position = "right")

print(Visualization3)


#V3cap <- fig_nums(name = "Visual3", caption = "Top three most studied organisms and the number of studies which proved lethal for that organism. 'Lethal' includes LD50 (50% morality) and NR-LETH (100% mortality), 'Not Lethal' includes measures with minimal or no effect on the organism.")
```






```{r visualization4, echo=FALSE}

Visualization4 <- ggplot(Three_dat_mutateMortality, aes(Mortality.Rate,))+
  geom_bar(stat = "count", position = "stack", aes(fill=Drug.Class))+
  labs(x="Effect on Mortality", y="Number of Studies", fill="Drug Class")+
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "gray72"))+
  theme(legend.position = "right")

print(Visualization4)

#V4cap <- fig_nums(name = "Visual4", caption = "Number of studies associated with each drug class and whether or not they proved lethal. Again, 'Lethal' includes LD50 (50% morality) and NR-LETH (100% mortality), 'Not Lethal' includes measures with minimal or no effect on the organism.")

```




```{r visualization5, echo=FALSE}
Visualization5<- ggplot(Three_dat_mutateMortality, aes(Endpoint,))+
  geom_bar(stat = "count", aes(fill= Drug.Class))+
  labs(x="Endpoint", y="Number of Studies", fill="Drug Class")+
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "gray72"))+
  facet_grid(Drug.Class~.)+
   theme(strip.text.y = element_text(angle = 0), legend.position = "none")
  
print(Visualization5)
  
#V5cap <- fig_nums(name = "Visual5", caption = "Endpoint effects of each drug class dependent on how many studies came to the same conclusion.")
  
```


<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, rationale for your approach, and the justification of meeting or failing to meet assumptions of tests.>


\newpage

# Summary and Conclusions
<Summarize your major findings from your analyses. What conclusions do you draw from your findings? Make sure to apply this to a broader application for the research question you have answered.>

